Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-30 survey/survey.tex
a bit more on sec 3

This discussion is purely ``semantic'', in the sense that it is about
the denotation of simple primitive data (sets) and their reversible
deformations (bijections).  We would like to reverse engineer a programming
language from this denotation. But first, an obvious remark: any two sets
$C$ and $D$ of cardinality $n$ are always in bijective correspondance. So
we can abstract away from the details of the elements of $C$ and $D$ and
instead choose canonical representations -- in much the same way as computers
choose binary words to represent everything.

@@

We now turn into the mathematical formalization of this idea. First we
note that the type structure of types has a nice correspondance
(Curry-Howard) to logic:

\begin{center}
\begin{tabular}{c|c}
Logic & Types \\ \hline
$\true$ & $\top$ \\
$\false$ & $\bot$ \\
$\land$ & $\times$ \\
$\lor$ & $\uplus$ \\
\end{tabular}
\end{center}

This correspondance is rather fruitful. As logical expressions form
a commutative semiring, we would expect that types too form a commutative
semiring. And indeed they do -- at least up to \emph{type equivalence}.  The
natural numbers $\Nat$ are another commutative semiring; it will turn out that,
even though the Curry-Howard correspondance has been extremely fruitful for
programming language research, it is $\Nat$ which will be a better model for
finite structured types.

@@

We cannot immediately use the above definition in a typed setting.
First, types do not naturally want to be put together into a ``set''.
This can be fixed if we replace the set $R$ with a universe $U$, and
rather than insisting that we have $0 \in R$, we have a typing
judgement $\bot : U$ (and similarly for the other items).

Our first instinct would be to similarly replace $=$ with
$\equiv$, i.e. propositional equality. This bring the second
problem to the fore: this is false. For example, unless $A \equiv B$,
$A \times B \equiv B \times A$ is not normally provable%
\footnote{Not without assuming that $U$ is univalent, which we
don't.}. But it should be clear that $A \times B$ and $B \times A$
contain equivalent information. In other words, we would like to
be able to witness that $A \times B$ can be reversibly deformed
into $B \times A$. Which motivates the introduction of
type \emph{equivalences}. To do this, we need a few important
auxiliary concepts first.

@@

If we replace $=$ by $\simeq$, then we can indeed prove that
types (with $\bot, \top, \uplus, \times$) form a commutative semiring.

@@

If we revisit the Curry-Howard correspondance, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow
  \qquad A \ \text{is inhabited} \]
The above is the real \textit{essence} of the correspondance.
In other words, classical Curry-Howard tells us about
\emph{logical equivalence} of types. This is even a
constructive statement: there are indeed functions
$f : A \uplus A \rightarrow A$ and $g : A \rightarrow A \uplus A$;
however, they are not inverses.

@@

Let us thus analyze the crux of the ``problem''. In logic, we have
that $\land$ and $\lor$ are both \emph{idempotent}: this is the
property of any binary operation $*$ where
$\forall a. a * a = a$. And it should be clear that idempotent
operations are \emph{forgetful}: its input has two copies of $a$,
but its output, only one. On the type side, something more subtle
happens. Consider $\top \uplus \top$ versus $\top$; the first has
\emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
while the second only one ($\texttt{tt}$). These cannot be put in
bijective correspondance. Even though the ``payload'' \texttt{tt}
is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
away information -- something we have expressly disallowed.
Yes, this should remind you of Maxwell's demon: even though the
data is the same, they are tagged differently, and this
data is indeed information, and must be preserved.

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& \tau & \simeq & \tau &: \idc \\
\\
\identlp :&  \bot \uplus \tau & \simeq & \tau &: \identrp \\
\swapp :&  \tau_1 \uplus \tau_2 & \simeq & \tau_2 \uplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \uplus (\tau_2 \uplus \tau_3) & \simeq & (\tau_1 \uplus \tau_2) \uplus \tau_3 &: \assocrp \\
\\
\identlt :&  \top \times \tau & \simeq & \tau &: \identrt \\
\swapt :&  \tau_1 \times \tau_2 & \simeq & \tau_2 \times \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \times (\tau_2 \times \tau_3) & \simeq & (\tau_1 \times \tau_2) \times \tau_3 &: \assocrt \\
\\
\distz :&~ \bot \times \tau & \simeq & \bot ~ &: \factorzl \\
\dist :&~ (\tau_1 \uplus \tau_2) \times \tau_3 & \simeq & (\tau_1 \times \tau_3) \uplus (\tau_2 \times \tau_3)~ &: \factor
\end{array}
\]
\caption{Type isomorphisms.}
\label{type-isos}
\end{figure}

Nevertheless, the Curry-Howard correspondance still has some
force. We know that the inhabitants of types formed with
with $\bot, \top, \uplus, \times$ are a commutative semiring. What we
want to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal?
We have a set of generators for those equations, namely those
in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
of mutually inverse functions which witness these identities.
For concreteness, we show the signatures in Figure~\ref{type-isos}.
As these all come in symmetric pairs (some of which are self-symmetric),
we give names for both directions. The names use $\iso$, $+$ and $*$
as these will be what we use when we reflect these isomorphism into
a programming language, which we proceed to do next.

@@

We now have in our hands our desired denotational semantics: we want
to create a language $\Pi$ such that the types and type combinators
map to $\bot, \top, \uplus, \times$, and such that we have ground
terms whose denotation are all $16$ type isomorphisms of
Figure~\ref{type-isos}. This is completely straightforward, as
we can simply do this literally. To make the analogy with
commutative semirings stand out even more, we will use $0, 1, +$, and $*$
at the type level, and will denote ``equivalence'' by $\iso$.
Thus Figure~\ref{pi-terms} shows the ``constants'' of the language.

@@

\amr{We should also have a quick survey of other works in reversible
computation? Most of the other work tries to start from well known
models of computation or well known programming languages, and then
adapts them to be reversible. Pi is different in that it starts from
the semantic implications of reversibility. Then, because it adds
types, rather than control-flow (or even data-flow) as its next layer
of ``understanding'', this leads to equivalences, isomorphisms, etc.
This then leads, quite naturally, to finding that the ``proof
language'' of semirings (and Rig Groupoids at level 2) is actually a
programming language. And it is Pi. This is a neat twist on
Curry-Howard because CH is about \textbf{inhabitation} only. But with
``conservation of information'' as the basis, a different kind of
correspondance arises; in fact, this one may well be an actual
isomorphism.}

>>>>

This discussion is purely ``semantic'', in the sense that it is about
the denotation of simple primitive data (sets) and their reversible
deformations (bijections).  We would like to reverse engineer a programming
language from this denotation. But first, an obvious remark: any two sets
$C$ and $D$ of cardinality $n$ are always in bijective correspondence. So
we can abstract away from the details of the elements of $C$ and $D$ and
instead choose canonical representations -- in much the same way as computers
choose binary words to represent everything.

@@

We now turn into the mathematical formalization of this idea. Our goal
is a sound and complete equivalence relation on types that maintains
the number of points. First we note that the type structure of types
has a nice correspondence (Curry-Howard) to logic:

\begin{center}
\begin{tabular}{c|c}
Logic & Types \\ \hline
$\false$ & $\mathbf{0}$ \\
$\true$ & $\mathbf{1}$ \\
$\land$ & $\times$ \\
$\lor$ & $\uplus$ \\
\end{tabular}
\end{center}

This correspondence is rather fruitful. As logical expressions form a
commutative semiring, we would expect that types too form a
commutative semiring. And indeed they do -- at least up to \emph{type
  isomorphism}.  The natural numbers $\Nat$ are another commutative
semiring; it will turn out that, even though the Curry-Howard
correspondence has been extremely fruitful for programming language
research, it is $\Nat$ which will be a better model for finite
structured types as the corresponding commutative semiring captures
the familiar numerical identities that evidently preserve the number
of points in the types.

@@

We would like to adapt the commutative semiring definition to the
setting of structured types. First, types do not naturally want to be
put together into a ``set''.  This can be fixed if we replace the set
$R$ with a universe $U$, and replace the set membership $0 \in R$ with
the typing judgement $\mathbf{0} : U$ (and similarly for the other
items). Our first instinct would be to similarly replace $=$ with a
judgement $A \equiv B$ that asserts that $A$ and $B$ are
\emph{propositionally equal}. This is however not true: unless
$A \equiv B$, $A \times B \equiv B \times A$ is not normally
provable\footnote{Not without assuming that $U$ is univalent, which we
  don't.}. But it should be clear that $A \times B$ and $B \times A$
contain equivalent information. In other words, we would like to be
able to witness that $A \times B$ can be reversibly deformed into
$B \times A$. Which motivates the introduction of type
\emph{equivalences}. To do this, we need a few important auxiliary
concepts first.

@@

If we replace $=$ by $\simeq$, then we can indeed prove that
types (with $\mathbf{0}, \mathbf{1}, \uplus, \times$) form a commutative semiring.

@@

If we revisit the Curry-Howard correspondence, we notice one
more issue. In logic, it is true that $A \lor A = A$ and
$A \land A = A$. However, neither $A \uplus A$ nor $A \times A$ are
equivalent to $A$. They are however \emph{equi-inhabited}. This is
a fancy way of saying
\[ A \uplus A \ \text{is inhabited} \qquad \Leftrightarrow
  \qquad A \ \text{is inhabited} \]
The above is the real \textit{essence} of the correspondence.
In other words, classical Curry-Howard tells us about
\emph{logical equivalence} of types. This is even a
constructive statement: there are indeed functions
$f : A \uplus A \rightarrow A$ and $g : A \rightarrow A \uplus A$;
however, they are not inverses.

@@

Let us thus analyze the crux of the ``problem''. In logic, we have
that $\land$ and $\lor$ are both \emph{idempotent}: this is the
property of any binary operation $*$ where
$\forall a. a * a = a$. And it should be clear that idempotent
operations are \emph{forgetful}: its input has two copies of $a$,
but its output, only one. On the type side, something more subtle
happens. Consider $\mathbf{1} \uplus \mathbf{1}$ versus $\mathbf{1}$; the first has
\emph{two} proofs of inhabitation ($\texttt{inl tt} and \texttt{inr tt}$)
while the second only one ($\texttt{tt}$). These cannot be put in
bijective correspondence. Even though the ``payload'' \texttt{tt}
is the same, forgetting $\texttt{inl}$ (or \texttt{inr}) throws
away information -- something we have expressly disallowed.
Yes, this should remind you of Maxwell's demon: even though the
data is the same, they are tagged differently, and this
data is indeed information, and must be preserved.

\begin{figure}[t]
\[
\begin{array}{rrcll}
\idc :& \tau & \simeq & \tau &: \idc \\
\\
\identlp :&  \mathbf{0} \uplus \tau & \simeq & \tau &: \identrp \\
\swapp :&  \tau_1 \uplus \tau_2 & \simeq & \tau_2 \uplus \tau_1 &: \swapp \\
\assoclp :&  \tau_1 \uplus (\tau_2 \uplus \tau_3) & \simeq & (\tau_1 \uplus \tau_2) \uplus \tau_3 &: \assocrp \\
\\
\identlt :&  \mathbf{1} \times \tau & \simeq & \tau &: \identrt \\
\swapt :&  \tau_1 \times \tau_2 & \simeq & \tau_2 \times \tau_1 &: \swapt \\
\assoclt :&  \tau_1 \times (\tau_2 \times \tau_3) & \simeq & (\tau_1 \times \tau_2) \times \tau_3 &: \assocrt \\
\\
\distz :&~ \mathbf{0} \times \tau & \simeq & \mathbf{0} ~ &: \factorzl \\
\dist :&~ (\tau_1 \uplus \tau_2) \times \tau_3 & \simeq & (\tau_1 \times \tau_3) \uplus (\tau_2 \times \tau_3)~ &: \factor
\end{array}
\]
\caption{Type isomorphisms.}
\label{type-isos}
\end{figure}

Nevertheless, the Curry-Howard correspondence still has some
force. We know that the inhabitants of types formed with
with $\mathbf{0}, \mathbf{1}, \uplus, \times$ are a commutative semiring. What we
want to know is, which types are equivalent? From a commutative semiring
perspective, this amounts to asking what terms are equal?
We have a set of generators for those equations, namely those
in~(\ref{eq:csemiring}). What we thus need is to create $8$ pairs
of mutually inverse functions which witness these identities.
For concreteness, we show the signatures in Figure~\ref{type-isos}.
As these all come in symmetric pairs (some of which are self-symmetric),
we give names for both directions. The names use $\iso$, $+$ and $*$
as these will be what we use when we reflect these isomorphism into
a programming language, which we proceed to do next.

@@

We now have in our hands our desired denotational semantics: we want
to create a language $\Pi$ such that the types and type combinators
map to $\mathbf{0}, \mathbf{1}, \uplus, \times$, and such that we have ground
terms whose denotation are all $16$ type isomorphisms of
Figure~\ref{type-isos}. This is completely straightforward, as
we can simply do this literally. To make the analogy with
commutative semirings stand out even more, we will use $0, 1, +$, and $*$
at the type level, and will denote ``equivalence'' by $\iso$.
Thus Figure~\ref{pi-terms} shows the ``constants'' of the language.

@@

\amr{We should also have a quick survey of other works in reversible
computation? Most of the other work tries to start from well known
models of computation or well known programming languages, and then
adapts them to be reversible. Pi is different in that it starts from
the semantic implications of reversibility. Then, because it adds
types, rather than control-flow (or even data-flow) as its next layer
of ``understanding'', this leads to equivalences, isomorphisms, etc.
This then leads, quite naturally, to finding that the ``proof
language'' of semirings (and Rig Groupoids at level 2) is actually a
programming language. And it is Pi. This is a neat twist on
Curry-Howard because CH is about \textbf{inhabitation} only. But with
``conservation of information'' as the basis, a different kind of
correspondence arises; in fact, this one may well be an actual
isomorphism.}
